Comparing probabilistic accounts of probability judgments

### 2021-11-07 paper-rmd/paper-papaja.Rmd
submitted

authornote: |
 This manuscript has not yet been peer-reviewed. This study was not preregistered.

abstract: |
 Bayesian theories of cognitive science hold that cognition is fundamentally probabilistic, but one mark sometimes held against these theories is that people’s explicit probability judgments often violate the laws of probability. Two recent proposals, the "Probability Theory plus Noise" [@costello.watts2014] and "Bayesian Sampler" [@zhu.etal2020] theories of probability judgments, both seek to account for these biases while maintaining that mental credences are fundamentally probabilistic. These theories fit quite differently into the larger project of Bayesian cognitive science, but their many similarities complicate comparisons of their predictive accuracy. In particular, comparing the models demands a careful accounting of model complexity. Here, I cast these theories into a Bayesian data analysis framework that supports principled model comparison using information criteria. Comparing the fits of both models on data collected by Zhu and colleagues [-@zhu.etal2020] I find the data are best explained by a modified version of the Bayesian Sampler model under which people may hold informative priors about probabilities.

@@

keywords          : "probability judgments, Bayesian cognitive science, heuristics and biases"
wordcount         : "4617"

@@

Bayesian theories of cognition offer a unified formal framework for cognitive science [@tenenbaum.etal2011] that has had remarkable explanatory successes across domains, including in perception [e.g. @kersten.etal2004], memory [e.g. @anderson1991], language [e.g. @xu.tenenbaum2007], and reasoning [e.g. @lu.etal2012]. At the heart of the Bayesian project is the idea that cognition is fundamentally probabilistic: that people reason according to subjective degrees of belief which follow the laws of probability and, in particular, that they are revised in light of evidence according to Bayes' Rule. It is somewhat embarrassing then, that these theories have often been accused of failing to describe human "beliefs" of the simple and everyday sort, such as beliefs like "it will rain tomorrow", "vaccines are safe," or "this politician is trustworthy".

@@

Costello & Watts [-@costello.watts2014; -@costello.watts2016; -@costello.watts2018] have proposed a theory of probability judgment they call the "Probability Theory plus Noise" theory (PT+N). In the PT+N model, mental "samples" are drawn from a probabilistic mental model of events and are then "read" with noise, so that some positive examples will be read as negative and some negative examples read as positive with some probability $d$. The results are probability judgments reflecting probabilistic credences perturbed by noise. In their model, the probability a mental sample for an event $A$ is read as A is the probability that the sample truly is $A$, $p(A)$, and that it is correctly read $(1-d)$, plus the probability that the sample is not $A$, $1-P(A)$ and that it is incorrectly read ($d$), or:

@@

Zhu, Sanborn, and Chater [-@zhu.etal2020] compared their Bayesian Sampler model against Costello & Watts’ (2014; 2016; 2017; 2018) PT+N model as explanations for human probability judgments in two experiments. Unfortunately, their results were somewhat equivocal. They fit both "simple" and "complex" versions of each model, where the complex versions of these models introduce additional parameters $d'$ and $N'$ that allow for different patterns of judgments for conjunctive and disjunctive judgments as compared with simple probability judgments. These additional parameters are crucial to both models’ explanations of conjunction fallacies---key findings in the probability judgment literature [@costello.watts2017; @zhu.etal2020]. They compared the fits of these models to data via Bayesian Information Criteria score.  Table 1 below presents the total BIC scores computed for each model as originally fit, using the authors’ original code and saved model outputs [@zhu.etal2020, Supplementary materials].

@@

In both experiments, the simple version of the Bayesian Sampler scores best, but the complex version of the PT+N model comes in second (lower BIC scores are better). It is not obvious what conclusions should be drawn from these results. Of course, the simple Bayesian Sampler model appears to win by the numbers. But conjunction fallacies are an extremely robust empirical finding [@mellers.etal2001; @sides.etal2002] and are clearly present in the data collected by Zhu and colleagues [-@zhu.etal2020]. So we might justifiably rule out the simple variants of the models on the grounds that they will fail to capture important qualitative features of the data, which could instead favor the PT+N theory. For their part, Zhu and colleagues [-@zhu.etal2020] chose to average the simple and complex model scores together. This approach somewhat favors the Bayesian Sampler theory overall, though they are cautious about drawing any strong conclusions in favor of their theory.

@@

There are at least three challenges to accounting for model complexity in the comparison of the PT+N and Bayesian Sampler models. First, the models differ not only in the number of parameters but in the domain of those parameters. Zhu and colleagues [-@zhu.etal2020] assume that the Bayesian Sampler model’s prior distribution should reflect ignorance or a lack of information. A uniform prior, Beta(1, 1) is the most obvious choice in this case, but theoretical arguments can also be made for Beta(0.5, 0.5) (Jeffrey’s prior), Beta(0,0) (Haldene’s prior), or perhaps other symmetric beta distributions Beta($\beta$,$\beta$) with $\beta \in [0, 1]$ [@jaynes2003]. Via the bridging conditions, they show that assuming $\beta \in [0, 1]$ restricts the "noise level" for the Bayesian Sampler, represented in implied $d$ under the PT+N model, to [0, $1/3$] (approaching $1/3$ as $N$ approaches 1), whereas the PT+N model permits noise values between $[0, 1/2]$.

@@

In addition, the Bayesian framework also supports straightforward implementation of hierarchical versions of these models with partial pooling. This allows for information about model parameters to be shared across participants, resulting in a potential improvements out-of-sample prediction, reductions in model complexity, and a more realistic test of the models.

@@

Zhu, Sanborn & Chater (2020) implement completely unpooled models with separate $d$, $d'$, $N$, $N'$, and $\beta$ parameters for each participant. Figure 1 displays the translation of the PT+N model into the Bayesian framework, along with a plate diagram representing the dependencies among parameters.

@@

Finally, I also explored fitting versions of the Bayesian Sampler model that allowed values of $\beta > 1$. Restricting $\beta$ to [0,1] restricts the prior distribution of the Bayesian sampler to the class of "ignorance priors" [@zhu.etal2020]. However, it is also possible that people bring informative priors to the probability judgment task. Indeed, Zhu and colleagues (2020) acknowledge there are situations where an informative prior may be warranted [see e.g., @fennell.baddeley2012]. If $\beta$ is unrestricted, allowed to fall in the domain $[0, \infty]$ then the Bayesian Sampler model becomes more flexible, allowing for equivalent "noise" levels in the same $[0, 1/2]$ range as the PT+N model. That is, through the bridging condition, the implied $d$ approaches $1/2$ in the limit as $N \to 1$ and $\beta \to \infty$. Though it would seem a more fundamental change, this same model may also be seen as a version of the PT+N theory that jettisons its constructive account of conditional probability judgment. Thus, fitting this additional unrestricted model allows for a complete comparison of the models along both of their differing dimensions.

@@

I fit each of the models specified above to data from Zhu et al’s [-@zhu.etal2020] Experiment 1 and 2 and estimated the expected log predictive density with PSIS-LOO ($\widehat{\text{elpd}}_{\text{LOO}}$) for each combination. Compared with BIC, $\widehat{\text{elpd}}_{\text{LOO}}$ offers a more sophisticated estimation of model complexity and is more appropriate in the "$\mathcal{M}$-open" case; situations where we do not know if any of the models being compared are the "true" model [@vehtari.etal2019]. Model posteriors were estimated using the Numpyro [@phan.etal2019] implementation of the No-U-Turn Hamiltonian Markov chain Monte Carlo (MCMC) sampler. For each model, four MCMC chains of 2000 iterations were sampled after 2000 iterations of warmup and all passed convergence tests according to $\hat{R}$ [see @gelman.etal2014a]. Figure 3 below displays the estimated differences in $\widehat{\text{elpd}}_{\text{LOO}}$ scores for each of the models as compared to the best-scoring model.

@@

Recall, the unrestricted Bayesian Sampler model may also be seen as a version of the PT+N theory excising its constructive account of conditional probability judgments. So, comparing the PT+N models to the unrestricted Bayesian Sampler model offers a test of the PT+N's constructive account of conditional probability jugments. Comparing predictions from the unrestricted Bayesian Sampler model and the best-fitting PT+N model, we see that the Bayesian Sampler model better captures these judgments from both experiments (see Table 2). And comparing their predictions specifically for conditional probability judgments, the unrestricted Bayesian Sampler again provides a better fit, especially for the modestly-correlated events of Experiment 2 (Exp 1: response-level $r$ = .91 vs. .89 (_p_ < .05); query-level $r$ = .98 vs. .97; Exp 2: response-level $r$ = .72 vs. .67 (_p_ < .01); query-level $r$ = .92 vs. .85) These findings suggest that conditioning is better seen as part of a mental model of the events than as part of the probability judgment process.

@@

Zhu and colleagues’ findings do cast some doubt on their proposal that the priors of the Bayesian Sampler model should reflect "ignorance priors," symmetric Beta distributions with $\beta \in [0, 1]$. As a generic prior that would be used across contexts, this class of uninformative priors has an appealing rational basis. Nevertheless, many participants' estimated $d$ and $d'$ parameters fell outside the $[0, 1/3]$ range implied by "ignorance" priors. These data suggest that people may instead bring informative priors to some probability judgment tasks. Given the different values of $d$ and $d'$ across Experiments 1 and 2, it seems likely that people bring domain-specific priors to judgment tasks, meaning that the most appropriate priors might be dictated by the context in which they make their judgments.

@@

One thing this model comparison has not decided, and likely *cannot* decide, is whether the distortions of probability judgments are products of mental noise or of further reasoning processes. Given the two models' tight connections via the bridging condition [@zhu.etal2020], it may not be possible to draw decisive conclusions here. Moreover, the theories may not be in any real competition over this point: Zhu and colleagues consider that "noise" might give an algorithmic-level solution to the computation-level task defined by the Bayesian Sampler [-@zhu.etal2020].

>>>>

authornote: |
 This manuscript has not yet been peer-reviewed. This study reports secondary data analyses and was not preregistered.

abstract: |
 Bayesian theories of cognitive science hold that cognition is fundamentally probabilistic, but people’s explicit probability judgments often violate the laws of probability. Two recent proposals, the "Probability Theory plus Noise" [@costello.watts2014] and "Bayesian Sampler" [@zhu.etal2020] theories of probability judgments, both seek to account for these biases while maintaining that mental credences are fundamentally probabilistic. These theories fit quite differently into the larger project of Bayesian cognitive science, but their many similarities complicate comparisons of their predictive accuracy. In particular, comparing the models demands a careful accounting of model complexity. Here, I cast these theories into a Bayesian data analysis framework that supports principled model comparison using information criteria. Comparing the fits of both models on data collected by Zhu and colleagues [-@zhu.etal2020] I find the data are best explained by a modified version of the Bayesian Sampler model under which people may hold informative priors about probabilities.

@@

keywords          : "probability judgments, Bayesian cognitive science, heuristics and biases"
wordcount         : "4826"

@@

Bayesian theories of cognition offer a unified formal framework for cognitive science [@tenenbaum.etal2011] that has had remarkable explanatory successes across domains, including in perception [e.g. @kersten.etal2004], memory [e.g. @anderson1991], language [e.g. @xu.tenenbaum2007], and reasoning [e.g. @lu.etal2012]. At the heart of the Bayesian project is the idea that cognition is fundamentally probabilistic: that people reason according to subjective degrees of belief which follow the laws of probability and, in particular, that they are revised in light of evidence according to Bayes' Rule. It is somewhat embarrassing then, that these theories have often been accused of failing to describe human "beliefs" of the simple and everyday sort, such as beliefs like "it will rain tomorrow", "vaccines are safe," or "this politician is trustworthy" [@chater.etal2020].

@@

Costello and Watts [-@costello.watts2014; -@costello.watts2016; -@costello.watts2018] have proposed a theory of probability judgment they call the "Probability Theory plus Noise" theory (PT+N). In the PT+N model, mental "samples" are drawn from a probabilistic mental model of events and are then "read" with noise, so that some positive examples will be read as negative and some negative examples read as positive with some probability $d$. The results are probability judgments reflecting probabilistic credences perturbed by noise. In their model, the probability a mental sample for an event $A$ is read as A is the probability that the sample truly is $A$, $p(A)$, and that it is correctly read $(1-d)$, plus the probability that the sample is not $A$, $1-P(A)$ and that it is incorrectly read ($d$), or:

@@

Zhu, Sanborn, and Chater [-@zhu.etal2020] compared their Bayesian Sampler model against Costello & Watts’ (2014; 2016; 2017; 2018) PT+N model as explanations for human probability judgments in two experiments. Unfortunately, their results were somewhat equivocal. They fit both "simple" and "complex" versions of each model, where the complex versions of these models introduce additional parameters $d'$ and $N'$ that allow for different patterns of judgments for conjunctive and disjunctive judgments as compared with simple probability judgments. These additional parameters are crucial to both models’ explanations of conjunction and disjunction fallacies---key findings in the probability judgment literature [@costello.watts2017; @zhu.etal2020]. They compared the fits of these models to data via Bayesian Information Criteria score.  Table 1 below presents the total BIC scores computed for each model as originally fit, using the authors’ original code and saved model outputs [@zhu.etal2020, Supplementary materials].

@@

In both experiments, the simple version of the Bayesian Sampler scores best, but the complex version of the PT+N model comes in second (lower BIC scores are better). It is not obvious what conclusions should be drawn from these results. Of course, the simple Bayesian Sampler model appears to win by the numbers. But conjunction and disjunction fallacies are extremely robust empirical findings [@mellers.etal2001; @sides.etal2002] and are clearly present in the data collected by Zhu and colleagues [-@zhu.etal2020]. So we might justifiably rule out the simple variants of the models on the grounds that they will fail to capture important qualitative features of the data, which could instead favor the PT+N theory. For their part, Zhu and colleagues [-@zhu.etal2020] chose to average the simple and complex model scores together. This approach somewhat favors the Bayesian Sampler theory overall, though they are cautious about drawing strong conclusions in favor of their theory.

@@

There are at least three challenges to accounting for model complexity in the comparison of the PT+N and Bayesian Sampler models. First, the models differ not only in the number of parameters but in the domain of those parameters. Zhu and colleagues [-@zhu.etal2020] assume that the Bayesian Sampler model’s prior distribution should reflect ignorance or a lack of information. A uniform prior, Beta(1, 1) is the most obvious choice in this case, but theoretical arguments can also be made for Beta(0.5, 0.5) (Jeffrey’s prior), Beta(0,0) (Haldene’s prior), or perhaps other symmetric beta distributions Beta($\beta$,$\beta$) with $\beta \in [0, 1]$ [@jaynes2003]. Via the bridging conditions, they show that assuming $\beta \in [0, 1]$ restricts the "noise level" for the Bayesian Sampler, represented in implied $d$ under the PT+N model, to fall within [0, $1/3$] (approaching $1/3$ as $N$ approaches 1), whereas the PT+N model permits noise values in $[0, 1/2]$.

@@

In addition, the Bayesian framework supports straightforward implementation of hierarchical versions of these models with partial pooling. This allows for information about model parameters to be shared across participants, resulting in potential improvements to out-of-sample prediction, reductions in model complexity, and a more realistic test of the models.

@@

Zhu, Sanborn & Chater (2020) implement completely unpooled models with separate $d$, $d'$, $N$, $N'$, and $\beta$ parameters for each participant. Although hierarchical models with partial pooling might be expected to better account for the data and offer a better test of the models, for consistency and comparison with Zhu et al.'s [@-zhu.etal2020] analyses, I first estimated implementations of these unpooled models. Figure 1 displays the translation of the PT+N model into the Bayesian framework, along with a plate diagram representing the dependencies among parameters.

@@

Finally, I also explored fitting a hierarchical version of the Bayesian Sampler model that allowed values of $\beta > 1$. Restricting $\beta$ to [0,1] restricts the prior distribution of the Bayesian sampler to the class of "ignorance priors" [@zhu.etal2020]. However, it is also possible that people bring informative priors to the probability judgment task. Indeed, Zhu and colleagues (2020) acknowledge there are situations where an informative prior may be warranted [see e.g., @fennell.baddeley2012]. If $\beta$ is unrestricted, allowed to fall in the domain $[0, \infty]$ then the Bayesian Sampler model becomes more flexible, allowing for equivalent "noise" levels in the same $[0, 1/2]$ range as the PT+N model. That is, through the bridging condition, the implied $d$ approaches $1/2$ in the limit as $N \to 1$ and $\beta \to \infty$. Though it would seem a more fundamental change, this same model may also be seen as a version of the PT+N theory that jettisons its constructive account of conditional probability judgment. Thus, fitting this additional unrestricted model allows for a complete comparison of the models along both of their differing dimensions.

@@

I fit each of the models specified above to data from Zhu et al’s [-@zhu.etal2020] Experiment 1 and 2 and estimated the expected log predictive density with PSIS-LOO ($\widehat{\text{elpd}}_{\text{LOO}}$) for each combination. Compared with BIC, $\widehat{\text{elpd}}_{\text{LOO}}$ offers a more sophisticated account of model complexity and is more appropriate in the "$\mathcal{M}$-open" case; situations where we do not know if any of the models being compared are the "true" model [@vehtari.etal2019]. Model posteriors were estimated using the Numpyro [@phan.etal2019] implementation of the No-U-Turn Hamiltonian Markov chain Monte Carlo (MCMC) sampler. For each model, four MCMC chains of 2000 iterations were sampled after 2000 iterations of warmup and all passed convergence tests according to $\hat{R}$ [see @gelman.etal2014a]. Figure 3 below displays the estimated differences in $\widehat{\text{elpd}}_{\text{LOO}}$ scores for each of the models as compared to the best-scoring model.

@@

Recall, the unrestricted Bayesian Sampler model may also be seen as a version of the PT+N theory that excises its constructive account of conditional probability judgments. So, comparing the PT+N models to the unrestricted Bayesian Sampler model offers a test of the PT+N's constructive account of conditional probability jugments. Comparing predictions from the unrestricted Bayesian Sampler model and the best-fitting PT+N model, we see that the Bayesian Sampler model better captures these judgments from both experiments (see Table 2). And comparing their predictions specifically for conditional probability judgments, the unrestricted Bayesian Sampler again provides a better fit, especially for the modestly-correlated events of Experiment 2 (Exp 1: response-level $r$ = .91 vs. .89 (_p_ < .05); query-level $r$ = .98 vs. .97; Exp 2: response-level $r$ = .72 vs. .67 (_p_ < .01); query-level $r$ = .92 vs. .85) These findings suggest that conditioning is better seen as part of a mental model of the events than as part of the probability judgment process.

@@

Zhu and colleagues’ findings do cast some doubt on their proposal that the priors of the Bayesian Sampler model should reflect "ignorance priors," symmetric Beta distributions with $\beta \in [0, 1]$. As a generic prior that would be used across contexts, this class of uninformative priors has an appealing rational basis. Nevertheless, many participants' estimated $d$ and $d'$ parameters fell outside the $[0, 1/3]$ range implied by "ignorance" priors. Given the different values of $d$ and $d'$ across Experiments 1 and 2, it seems likely that people bring informative, domain-specific priors to some judgment tasks, meaning that the most appropriate priors might be dictated by the context in which they make their judgments.

@@

One thing this model comparison has not decided, and likely *cannot* decide, is whether distortions of probability judgments are products of mental noise or of further reasoning processes. Given the two models' tight connections via the bridging condition [@zhu.etal2020], it may not be possible to draw decisive conclusions here. Moreover, the theories may not be in any real competition over this point: Zhu and colleagues consider that "noise" might give an algorithmic-level solution to the computation-level task defined by the Bayesian Sampler [-@zhu.etal2020].
